Commodity Flow Survey Exploration by Mike Gysel
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load packages
# Data visualization
library(ggplot2)
# Data Wrangling
library(dplyr)
# Pairplot
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
cfs <- read.csv('data/cfs_2012_pumf_csv.csv')
# Easier to work with sample
cfs_sample <- cfs[sample(nrow(cfs), 10000),]
```

# Summary of 2012 Commodity Flow Survey
The 2012 Commodity Flow Survey (CFS) is a joint effort by the Bureau of 
Transportation Statistics and the U.S. Census Bureau, U.S. Department of 
Commerce. The survey is the primary source of national and state-level data on 
domestic freight shipments by establishments in mining, manufacturing, whole 
sale auxiliaries, and selected retail and services trade industries, located in 
the 50 states and the District of Columbia. The survey produces estimates on 
the type, origin and destination, value, weight, modes of transportation, 
distance shipped, and ton-miles of commodities shipped. 

Additional information on the Commodity Flow Survey can be found at 
https://catalog.data.gov/dataset/commodity-flow-survey

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
# Data structure
str(cfs)
# Summary
summary(cfs)
```

There are 4,547,661 observations of  20 variables. 

Some interesting initial findings:

* The median shipment value is $875 and the mean shipment value is $18,280, 
which is dramatically higher due to the max shipment value of $521,277,800. 
What could have been shipped for over half a billion dollars?!!
* The median shipment weight is 270 pounds and the mean shipment weight is 
37,587 pounds, which is dramatically higher due to the maximum shipment weight 
of 277,029,600 pounds!! What could have weighed 277 million pounds?!
* The median shipment distance is 140 miles and the mean shipment distance is 
475 miles, with a maximum shipment distance of 6,932 miles. This shows that the 
majority of shipments were made in the US, with some made abroad, which takes 
us to
* 23 times as many shipments were not exported (4,361,940 shipments) as were 
exported (185,721 shipments).

# Univariate Analysis

## Origin State
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the number of occurrences of an origin state
ggplot(data=cfs, aes(x=ORIG_STATE)) + 
  geom_bar() +
  scale_x_continuous(breaks=seq(0,60,2))

# Determine origin states with most/least shipments
count.ORIG_STATE <- arrange(count(cfs, ORIG_STATE), desc(n))
```

The states with the most outgoing shipments are California, Texas, Ohio, 
Pennsylvania, and New York. The states with the least outgoing shipments are 
Alaska, Wyoming, Delaware, New Mexico, and Vermont.

I would like to check out a histogram of the number of shipments for each 
origin state.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram for number of outgoing shipments from an ORIG_STATE
ggplot(data=count.ORIG_STATE, aes(x=n)) + 
  geom_histogram(binwidth=20000) + 
  scale_x_continuous(breaks=seq(0,400000,50000))
```

The distribution of number of outgoing shipments from each state is a 
right-skewed, long tail distribution.

## Destination State
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the number of occurrences of a destination state
ggplot(data=cfs, aes(x=DEST_STATE)) + 
  geom_bar() + 
  scale_x_continuous(breaks=seq(0,60,2))

# Determine destination states with the most/least shipments
count.DEST_STATE <- arrange(count(cfs, DEST_STATE), desc(n))
```

The states with the most incoming shipments are California, Texas, New York, 
Florida, and Ohio. The states with the least incoming shipments are Rhode 
Island, Wyoming, Alaska, Vermont, and Delaware.

I would like to check out a histogram of the number of shipments for each 
destination state.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram for number of outgoing packages from a DEST_STATE
ggplot(data=count.DEST_STATE, aes(x=n)) + 
  geom_histogram(binwidth=15000) + 
  scale_x_continuous(breaks=seq(0,400000,50000))
```

The distribution of number of incoming shipments to each state, similar to outgoing shipments from each state, is a right-skewed, long tail distribution.

## Metro Area

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine the origin metro areas with the most shipments
count.ORIG_MA <- arrange(count(cfs, ORIG_MA), desc(n))

# Determine the destination metro areas with the most shipments
count.DEST_MA <- arrange(count(cfs, DEST_MA), desc(n))

count.ORIG_MA
count.DEST_MA
```

The top 5 origin and destination metro areas with the most outgoing/incoming 
shipments are the same: New York-Newark, Los Angeles-Long Beach, 
Chicago-Naperville, Boston-Worcester-Providence, and 
Philadelphia-Reading-Camden.
Note: MA 99999 represents the remainder of numerous different areas, and was 
not counted here. 

I would like to check out a histogram of the number of shipments for each 
origin and destination metro area.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram for number of outgoing shipments from a ORIG_MA
ggplot(data=count.ORIG_MA, aes(x=n)) + 
  geom_histogram(binwidth=2000) + 
  scale_x_continuous(lim=c(0,150000))

# Histogram for number of outgoing shipments from a DEST_MA
ggplot(data=count.DEST_MA, aes(x=n)) + 
  geom_histogram(binwidth=5000) + 
  scale_x_continuous(lim=c(0,150000))
```

The distribution of number of incoming shipments to each metro area and the 
distribution of number of outgoing shipments from each metro area both display
a right-skewed, long tail distribution. This is what we would expect, given the
distribution of shipment quantity for each state.

## NAICS (North American Industry Classification System) Codes

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine the NAICS industry classifications with the most shipments
count.NAICS <- arrange(count(cfs, NAICS), desc(n))
count.NAICS
```

The industries with the most shipments are the Chemical Manufacturing Sector;
Aerospace Sector; Machinery, Equipment, and Supplies Merchant Wholesalers; Food 
Manufacturing; and Grocery and Related Product Merchant Wholesalers. It appears 
America likes to ship chemicals, machines, and food!!

I would also like to view the distribution of shipments per industry.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of shipments in each industry
ggplot(data=count.NAICS, aes(x=n)) + 
  geom_histogram(binwidth=5000)
```

The distribution of shipments for each industry appears relatively symmetric, 
with a bit of a tail to the right.

## SCTG (Standard Classification of Transported Goods) Codes

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine the SCTG with the most shipments
count.SCTG <- arrange(count(cfs, SCTG), desc(n))
count.SCTG
```

The industries with the most shipments are Electronic and Other Electrical 
Equipment and Components, and Office Equipment; Plastics and Rubber; 
Mixed Freight; Machinery; and Miscellaneous Manufactured Products. These 
largely align with NAICS Codes, with Plastics and Rubber aligning with the 
Chemical Manufacturing Sector, and both Electrical Equipment and Machinery
aligning with Machinery.

## Quarter

```{r}
# Determine the number of shipments for each quarter
arrange(count(cfs, QUARTER), desc(n))
```

The distribution of shipments is relatively even, with slightly fewer shipments
each subsequent quarter.

## Mode

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine the mode with the most shipments
count.MODE <- arrange(count(cfs, MODE), desc(n))

# Bar graph of each mode
ggplot(data=cfs, aes(x=MODE)) + 
  geom_bar()
```

The three modes of transit for each shipment with by far the most shipments
are Private Truck; For-hire Truck; and Parcel, USPS, or Courier. This makes me 
wonder which modes are typically used for different attributes, such as 
distance, shipment value, shipment type, etc.

## Shipment Value

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Modify cfs$SHIPMT_VALUE numeric
cfs$SHIPMT_VALUE <- as.numeric(cfs$SHIPMT_VALUE)

# Plot a histogram of shipment value
ggplot(data=cfs, aes(SHIPMT_VALUE)) + 
  geom_histogram(binwidth=1000) + 
  scale_x_continuous(lim=c(0,10000000)) +
  scale_y_continuous(lim=c(0,100))
```

The distribution of shipment value is a right-skewed, long tail distribution. 
This distribution makes me want to use a logarithmic scale on the x-axis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot a histogram of shipment value, x-axis logarithmic
ggplot(data=cfs, aes(SHIPMT_VALUE)) + 
  geom_histogram() + 
  scale_x_log10()
```

As expected, the logarithmic transformed shipment value distribution is 
relatively symmetric.

## Shipment Weight
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Modify cfs$SHIPMT_WGHT numeric
cfs$SHIPMT_WGHT <- as.numeric(cfs$SHIPMT_WGHT)

# Plot a histogram of shipment value
ggplot(data=cfs, aes(SHIPMT_WGHT)) + 
  geom_histogram(binwidth=1000) + 
  scale_x_continuous(lim=c(0,10000000)) +
  scale_y_continuous(lim=c(0,100))
```

The distribution of shipment weight is right-skewed with a peak less than 100 
pounds, and a secondary peak around 3 million pounds. This distribution makes 
me want to use a logarithmic scale on the x-axis. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot a histogram of shipment weight, x-axis logarithmic
ggplot(data=cfs, aes(SHIPMT_WGHT)) + 
  geom_histogram(binwidth=.2) + 
  scale_x_log10()
```

This logarithmic transformed shipment value distribution looks relatively flat 
with a low and high peak.

## Shipment Distance Routed

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Modify cfs$SHIPMT_DIST_ROUTED numeric
cfs$SHIPMT_DIST_ROUTED <- as.numeric(cfs$SHIPMT_DIST_ROUTED)

# Plot a histogram of shipment distance
ggplot(data=cfs, aes(SHIPMT_DIST_ROUTED)) + 
  geom_histogram(binwidth=100)
```

The distribution of shipment distance routed is a right-skewed, long tail 
distribution. This distribution makes me want to use a logarithmic scale on the 
x-axis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=cfs, aes(SHIPMT_DIST_ROUTED)) + 
  geom_histogram(binwidth=.05) + 
  scale_x_log10()
```

Unexpectedly, the logarithmic distribution is bimodal, with a peak around 10 
pounds and a larger peak around 1,000 pounds. This makes me wonder what 
different attributes these two peaks have.

## Export Country

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Look at the exports by category
# Bar Chart
ggplot(data=cfs, aes(x=EXPORT_CNTRY)) +
  geom_bar()
# Table
arrange(count(cfs, EXPORT_CNTRY), desc(n))
```
As expected, most (95.9%) shipments were not exports. Of the shipments that 
were exports, 29.4% went to Canada, 10.4% went to Mexico, and 60.2% wentto 
other countries.

## Hazmat

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Table of distribution of hazardous vs. non-hazardous shipments
count(cfs, HAZMAT)
```

93.3% of shipments are Not Hazmat (good to hear), 2.9% are Class 3.0 Hazmat 
(flammable liquids), and 3.7% are Other Hazmat.

### What is the structure of your dataset?

There are 4,547,661 shipments with 20 characteristics (shipment identifier, 
origin state, origin metro area, origin CFS area, destination state, 
destination metro area, destination CFS area, NAICS industry classification, 
quarter shipment occurred, SCTG commodity code, mode of transportation, 
shipment value, shipment weight, great circle distance, routed distance, 
temperature control, export shipment, export destination, hazardous material, 
and weighting factor).

### What is/are the main feature(s) of interest in your dataset?

Some primary inquiries I have are:

* What shipments weigh and cost so much?
* How does mode of transportation vary with shipment distance and weight?
* What are the differences in great circle distance and routed distance. How do 
these vary as distance increases? Does the ratio of great circle distance to 
routed distance increase with exports?
* Shipment weights peak around 10 pounds and 1,000 pounds. What are the 
attributes of these two peaks?
* What are attributes of packages we send abroad?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think most of the features can potentially be useful with this broad range of 
inquiries above. 

### Did you create any new variables from existing variables in the dataset?

I counted the occurrences of different categorical variables, but did not 
create any new ones.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Per above, the distribution of shipment weights, with peaks around 10 pounds 
and 1,000 pounds is interesting.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
# Subset the data for a pairplot, the only relevant rows are
# QUARTER (9), SHIPMT_VALUE (12) SHIPMT_WGHT (13), SHIPMT_DIST_ROUTED (15), 
# TEMP_CNTL_YN (16) EXPORT_YN (17), HAZMAT (19)  
cfs_pairplot <- cfs_sample %>% select(SHIPMT_VALUE, 
                                      SHIPMT_WGHT, 
                                      SHIPMT_DIST_ROUTED, 
                                      TEMP_CNTL_YN, 
                                      EXPORT_YN, 
                                      HAZMAT)
ggpairs(cfs_pairplot, upper=list(continuous = wrap(ggally_points, size = .1)), 
        columnLabels = c("Value", 
                         "Weight", 
                         "Distance", 
                         "Temp", 
                         "Export", 
                         "Hazmat"))
```

Most variables show no relation, but some interesting patterns:

* Weak correlation between shipment weight and shipment value.
* Temperature controlled shipments are of greater value and weight on average.
* Export shipments are of greater value and weight on average.
* Hazmat shipment categories have different average values and weights.
* Distance routed differs by temperature control, export, and hazmat.

## Shipment Weights - What shipments weigh so much?

### Top 1% of Shipments by Weight

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine top 1% of shipment weights
summary(subset(cfs, SHIPMT_WGHT>quantile(SHIPMT_WGHT, 0.99)))
```
Some initial interesting findings:

* Most are not temperature controlled
* Most are not exported
* Most are not Hazmat

### What NAICS and SCTG Codes have shipments with the highest average weights?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# NAICS

# Determine shipment weight per shipment, average by NAICS Codes
arrange(cfs %>% group_by(NAICS) %>% summarise(avg=mean(SHIPMT_WGHT)), desc(avg))

# Determine the shipment weight per shipment, average by SCTG Codes
arrange(cfs %>% group_by(SCTG) %>% summarise(avg=mean(SHIPMT_WGHT)), desc(avg))
```
Some observations related to shipment weight:

* The average shipment weight for Mining and Quarrying (except oil and gas) at 
588,041 pounds, Farm Product Raw Material Merchant Wholesaler at 403,757 pounds, 
and Petroleum and Coal Product Manufacturing at 354,552 pounds far exceed the 
shipment weights for any other NAICS Codes. 
* The average shipment weight for Coal at 5,596,597 pounds and Metallic Ores 
and Concentrates at 1,683,167 pounds far exceed any other SCTG Codes.

### What transportation modes are used for shipments with the highest weights?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# What transportation mode do the top shipments by weight use?

# Top 1%
cfs.WEIGHT_1 <- 
  arrange(count(arrange(subset(cfs, 
                               SHIPMT_WGHT>quantile(SHIPMT_WGHT, 0.99)), 
                        SHIPMT_WGHT), MODE), desc(n))
# Find percentage of each mode
cfs.WEIGHT_1$perc <- cfs.WEIGHT_1$n/sum(cfs.WEIGHT_1$n)

# Top 0.1%
cfs.WEIGHT_P1 <- 
  arrange(count(subset(subset(cfs, SHIPMT_WGHT>quantile(SHIPMT_WGHT, 0.99)), SHIPMT_WGHT>quantile(SHIPMT_WGHT, 0.9)), MODE), desc(n))

# Find percentage of each mode
cfs.WEIGHT_P1$perc <- cfs.WEIGHT_P1$n/sum(cfs.WEIGHT_P1$n)

# Display
cfs.WEIGHT_1$perc
cfs.WEIGHT_P1$perc
```

Some observations related to shipment weight and transportation mode:

* For shipments in the top 1% of weight, 55% are transported by rail and 11% 
are transported by truck and rail.
* For shipments in the top 0.1% of weight, 54% are transported by rail, 16% are 
transported by pipeline, and 10% are transported by Inland Water.

This makes me wonder how heavy export shipments are transported.

### Are shipment weight and distance correlated?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Are shipment weight and shipment distance correlated?
ggplot(data=cfs_sample, aes(x=SHIPMT_WGHT, y=SHIPMT_DIST_ROUTED)) +
  geom_point(size=0.1, alpha=0.05) + 
  scale_x_continuous(lim=c(0,10000))

# Try log/quadratic models on the subsetted data
cfs_lm <- subset(cfs, SHIPMT_WGHT<10000)

lin.model<- lm(cfs_lm$SHIPMT_DIST_ROUTED ~ cfs_lm$SHIPMT_WGHT)
summary(lin.model)

log.model <- lm(cfs_lm$SHIPMT_DIST_ROUTED ~ cfs_lm$SHIPMT_WGHT + (log(cfs_lm$SHIPMT_WGHT)))
summary(log.model)

quadratic.model <- lm(cfs_lm$SHIPMT_DIST_ROUTED ~ cfs_lm$SHIPMT_WGHT + (cfs_lm$SHIPMT_WGHT^2))
summary(quadratic.model)
```

It appears there is an exponential relationship with shipment weight and routed
shipment distance, when removing outliers above a 10,000 pound shipment weight. 
The linear, exponential, and quadratic models however, with very low 
correlation coefficients, say otherwise.

### Are shipment weight and value correlated?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot shipment weight and shipment value
ggplot(data=cfs_sample, aes(x=SHIPMT_WGHT, y=SHIPMT_VALUE)) +
  geom_point(size=.1, alpha=0.5) + 
  scale_x_continuous(lim=c(0,100000)) + 
  scale_y_continuous(lim=c(0,1000000))

# Linear model of shipment weight and value
linear.model <- lm(cfs$SHIPMT_VALUE ~ cfs$SHIPMT_WGHT)
summary(linear.model)
```
The bulk of the data does not show much of a correlation between shipment 
weight and value. The R-Squared value of only 0.01 supports this. This is 
surprising, as I would have expected very heavy shipments to be worth a lot 
more than very light shipments.

### What is the median/average shipment weight for different exports?

```{r echo=FALSE, message=FALSE, warning=FALSE}
cfs.WEIGHT_EXPORT <- cfs %>% 
  group_by(EXPORT_CNTRY) %>% 
  summarise(median=median(SHIPMT_WGHT), avg=mean(SHIPMT_WGHT))
cfs.WEIGHT_EXPORT$median_perc <- cfs.WEIGHT_EXPORT$median/sum(cfs.WEIGHT_EXPORT$median)
cfs.WEIGHT_EXPORT$avg_perc <- cfs.WEIGHT_EXPORT$avg/sum(cfs.WEIGHT_EXPORT$avg)
cfs.WEIGHT_EXPORT$avg_perc
```

The median exports weight to Mexico dramatically exceeds the median export
weights anywhere else. What is being exported to Mexico?

```{r}
arrange(count(subset(cfs, EXPORT_CNTRY=='M'), NAICS), desc(n))
```

The NAICS Codes with the most shipments to Mexico are Fabricated Metal Product 
Manufacturing, Chemical Manufacturing, Transportation Equipment Manufacturing, 
Plastics and Rubber Products Manufacturing, and Machinery Manufacturing.

## Shipment Values - What shipments cost so much?

### Top 1% of Shipments by Value

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine top 1% of shipments by cost
summary(subset(cfs, SHIPMT_VALUE>quantile(SHIPMT_VALUE,0.99)))
```

Some initial interesting findings:

* Most are not temperature controlled
* Most are not exported
* Most are not Hazmat

### What NAICS and SCTG Codes have shipments with the highest average values?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# NAICS

# Determine shipment value per shipment, average by NAICS Codes
arrange(cfs %>% 
          group_by(NAICS) %>% 
          summarise(avg=mean(SHIPMT_VALUE)), desc(avg))

# Determine shipment value per shipment, average by 
arrange(cfs %>% 
          group_by(SCTG) %>% 
          summarise(avg=mean(SHIPMT_VALUE)), desc(avg))
```

Some observations related to shipment value:

* The average shipment value for Transportation Equipment Manufacturing at 
$192,388, Petroleum and Coal Product Manufacturing at $138,415, and Farm Product 
Raw Material Merchant Wholesalers at $81,859 far exceed the average shipment 
values for any other NAICS Code.
* The average shipment value for Transportation Equipment, not esewhere 
classified at $355,346; Metallic Ores and Concentrates at $145,027; and 
Gasoline, Aviation Turbine Fuel, and Ethanol (includes Kerosene, and Fuel 
Alcohols) have the greatest average shipment values among SCTG Codes.
It appears transportation shipments cost a lot!

### What transportation modes are used for shipments with the highest values?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# What transportation mode do the top shipments by value use?

# Top 1%
cfs.VALUE_1 <- 
  arrange(count(arrange(subset(cfs, 
                               SHIPMT_VALUE>quantile(SHIPMT_VALUE, 0.99)), 
                        SHIPMT_VALUE), MODE), desc(n))
# Find percentage of each mode
cfs.VALUE_1$perc <- cfs.VALUE_1$n/sum(cfs.VALUE_1$n)

# Top 0.1%
cfs.VALUE_P1 <- 
  arrange(count(subset(subset(cfs, SHIPMT_VALUE>quantile(SHIPMT_VALUE, 0.99)), SHIPMT_VALUE>quantile(SHIPMT_VALUE, 0.9)), MODE), desc(n))

# Find percentage of each mode
cfs.VALUE_P1$perc <- cfs.VALUE_P1$n/sum(cfs.VALUE_P1$n)

# Display
cfs.VALUE_1$perc
cfs.VALUE_P1$perc
```

Some observations related to shipment value and transportation mode:

* For shipments in the top 1% of value, 56% are transported by For-hire Truck, 
11% are transported by Rail, and another 11% are transported by Private Truck.
* For shipments in the top 0.1% of value, 41% are transported by For-Hire Truck 
and 22% are transported by Pipeline.

### What is the median/average shipment value for different exports?

```{r echo=FALSE, message=FALSE, warning=FALSE}
cfs.VALUE_EXPORT <- cfs %>% 
  group_by(EXPORT_CNTRY) %>% 
  summarise(median=median(SHIPMT_VALUE), avg=mean(SHIPMT_VALUE))

cfs.VALUE_EXPORT$median_perc <- 
  cfs.VALUE_EXPORT$median/sum(cfs.VALUE_EXPORT$median)
cfs.VALUE_EXPORT$avg_perc <- cfs.VALUE_EXPORT$avg/sum(cfs.VALUE_EXPORT$avg)

cfs.VALUE_EXPORT
```

The median exports value to each country is relatively similar, but the mean
exports value to Other countries (not Mexico or Canada) is significantly higher
than other exports. What NAICS/SCTG Codes are being exported to Other Countries, 
for exports with the highest values?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Find exports to Other countries with top 1% of Shipment values
cfs.VALUE_EXPORT_O1 <- 
  subset(cfs, EXPORT_CNTRY == 'O' & SHIPMT_VALUE>quantile(SHIPMT_VALUE, 0.99))

# NAICS
arrange(subset(cfs, EXPORT_CNTRY == 'O') %>% 
          group_by(NAICS) %>% 
          summarise(avg=mean(SHIPMT_VALUE)), desc(avg))

# SCTG
arrange(subset(cfs, EXPORT_CNTRY == 'O') %>% 
          group_by(SCTG) %>% 
          summarise(avg=mean(SHIPMT_VALUE)), desc(avg))
```

The industries with the most value shipped to Other countries are:

* NAICS Codes: Farm Product Raw Material Merchant Wholesalers, Transportation 
Equipment Manufacturing, and Petroleum and Coal Manufacturing.
* SCTG Categories: Fuel Oils (includes Diesel, Bunker C, and Biodiesel); 
Gasoline, Aviation Turbine Fuel, and Ethanol (includes Kerosene, and Fuel 
Alcohols); and Cereal Grains (includes seed).

The industries with the most value being shipped, regardless of how the data is 
subset, appear to all be related to Agriculture, Transportation, and Energy.

## What mode of transit is used for different distances?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine median/mean distance for each mode of transit
cfs.MODE_DIST <- cfs %>% 
  group_by(MODE) %>% 
  summarise(median=median(SHIPMT_DIST_ROUTED), mean=mean(SHIPMT_DIST_ROUTED))   

# Graph mean distance vs. mode of transit
ggplot(data=cfs.MODE_DIST, aes(x=MODE, y=mean)) +
  geom_col() + 
  scale_x_discrete()

# Graph median distance vs. mode of transit
ggplot(data=cfs.MODE_DIST, aes(x=MODE, y=median)) +
  geom_col() + 
  scale_x_discrete()
```

Air (incl truck & air), Truck and rail, and Truck and water are the three modes 
of transit with the highest mean shipment distance. Air (incl truck & air), 
and Truck and rail were the modes of transit with the highest median shipment 
distance.

## What are the differences in great circle distance and routed distance.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine ratio of great circle distance to routed distance
cfs$dist_ratio <- cfs$SHIPMT_DIST_ROUTED/cfs$SHIPMT_DIST_GC

# Summary of dist_ratio
summary(cfs$dist_ratio)
```

The median and mean ratio between great circle distance and routed distance are 
1.2128 and 1.3955, respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the dist_ratio and great circle distance
ggplot(data=cfs, aes(x=SHIPMT_DIST_GC, y=dist_ratio)) +
  geom_point()

# Try log/quadratic models on the subsetted data
lin.model<- lm(cfs$dist_ratio ~ cfs$SHIPMT_DIST_GC)
summary(lin.model)

log.model <- lm(cfs$dist_ratio ~ cfs$SHIPMT_DIST_GC + (log(cfs$SHIPMT_DIST_GC)))
summary(log.model)

quadratic.model <- lm(cfs$dist_ratio ~ cfs$SHIPMT_DIST_GC + (cfs$SHIPMT_DIST_GC^2))
summary(quadratic.model)
```

It appears there is an exponential relationship with  great circle
shipment distance and distance ratio. The linear, exponential, and quadratic 
models however, each with low correlation coefficients, say otherwise.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

There are few correlations between quantitative variables in the dataset, but 
the most interesting findings are:

* The shipments with the most weight and value tend to be in the agriculture, 
transportation, and energy industries.
* The heaviest shipments are transported mainly by rail, while the average 
shipments are transported mainly by road.
* Air, water, and rail are the modes of transit used for the longest shipment 
distances.
* The average weight of shipments exported to Mexico far exceed shipments
exported elsewhere; the average value of shipments exported to other countries
far exceed shipments exported to Mexico or Canada.

### What was the strongest relationship you found?

The correlation coefficient between shipment weight and shipment value was
0.33, per the pairplot.

# Multivariate Plots Section

## Shipment distance routed had a peak around 10 miles and a peak around 1000 \
miles. What are the different attributes of these peaks?
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Subset the data between (0,50) and (100,2000)
cfs.low_dist <- subset(cfs, SHIPMT_DIST_ROUTED<50) 
cfs.high_dist <- subset(cfs, SHIPMT_DIST_ROUTED>1000 & SHIPMT_DIST_ROUTED<2000)

# Look at summaries of first/second peaks
summary(cfs.low_dist)
summary(cfs.high_dist)
```

The low and  high peak for shipment distances appear to differ in shipment 
weight and shipment value.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create df of shipment weight counts for low_distance peak
cfs.low_dist_wght <- count(cfs.low_dist, SHIPMT_WGHT)
cfs.low_dist_wght$category <- 'low_dist'
# Create df of shipment weight counts for high_distance peak
cfs.high_dist_wght <- count(cfs.high_dist, SHIPMT_WGHT)
cfs.high_dist_wght$category <- 'high_dist'
# Combine low_dist and high_dist df's
cfs.dist_wght <- rbind(cfs.low_dist_wght, cfs.high_dist_wght)

# Histogram of shipment weight for the low and high distance peaks
ggplot(data=cfs.dist_wght, aes(x=SHIPMT_WGHT)) +
  geom_histogram(aes(color=category)) +
  scale_x_continuous(lim=c(0,300000))
```

The distribution of shipment weight for the low-distance (0 to 50 miles) and 
high-distance (1,000 to 2,000 miles) shipments appears to be very similar. It 
seems like this graph would be better if the count were a proportion of total 
count for each distance category, as this would account for the total number of 
shipments.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Determine ratio of count/total count
cfs.low_dist_wght$ratio <- cfs.low_dist_wght$n/sum(cfs.low_dist_wght$n) 
cfs.high_dist_wght$ratio <- cfs.high_dist_wght$n/sum(cfs.high_dist_wght$n)
# Combine low_dist and high_dist df's
cfs.dist_wght <- rbind(cfs.low_dist_wght, cfs.high_dist_wght)

# Line graph of shipment weight for the low and high distance peaks, 
# low shipment weight
ggplot(data=cfs.dist_wght, aes(x=SHIPMT_WGHT, y=ratio)) +
  geom_point(aes(color=category)) +
  scale_x_continuous(lim=c(0,50)) + 
  scale_y_continuous(lim=c(0,0.06)) +
  geom_smooth(aes(color=category))

# Line graph of shipment weight for the low and high distance peaks, 
# medium shipment weight
ggplot(data=cfs.dist_wght, aes(x=SHIPMT_WGHT, y=ratio)) +
  geom_point(aes(color=category)) +
  scale_x_continuous(lim=c(50,100)) + 
  scale_y_continuous(lim=c(0,0.004)) +
  geom_smooth(aes(color=category))

# Line graph of shipment weight for the low and high distance peaks,
# high shipment weight
ggplot(data=cfs.dist_wght, aes(x=SHIPMT_WGHT, y=ratio)) +
  geom_point(aes(color=category), size=.2) +
  scale_x_continuous(lim=c(100,5000)) + 
  scale_y_continuous(lim=c(0,0.0003))
```

These three line graphs show that a higher proportion of high distance 
shipments have low shipment weights. At approximately 65 pound shipment weight,
the proportion of low distance shipments surpasses that of high distance
shipments.

## dist_ratio for exports and non-exports

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the routed distance vs. dist_ratio (great circle distance/routed 
# distance) for non-exports and different types of exports
ggplot(data=cfs, aes(x=SHIPMT_DIST_GC, y=dist_ratio)) +
  geom_point(aes(color=cfs$EXPORT_CNTRY), size=0.1) +
  scale_x_continuous(lim=c(0,100))
```

The patterns of distance ratio behave pretty erratically. The only patterns
that can be observed are the ratio decreasing as distance increases. The 
distance ratio for non Mexico/Canada exports also appear to be the highest 
of any group.

## How does mode of transit change with shipment distance?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Change MODE column from int to factor
cfs$MODE <- as.factor(cfs$MODE)

# Plot the percentages of packages transported via different modes,
# by shipment distance
ggplot(data=cfs, aes(x=SHIPMT_DIST_ROUTED)) +
  geom_freqpoly(aes(y=..count../sum(..count..), color=MODE)) + 
  scale_x_continuous(breaks=seq(0,6000,500), lim=c(0,6000)) +
  scale_y_continuous(labels=scales::percent)
```

Now subset the data to only include the predominant modes of transport.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the percentages of packages transported by predominant modes
ggplot(data=subset(cfs, MODE==04 | MODE==05 | MODE==14), aes(x=SHIPMT_DIST_ROUTED)) +
  geom_freqpoly(aes(y=(..count..)/sum(..count..), color=MODE)) + 
  scale_x_continuous(breaks=seq(0,6000,500), lim=c(0,6000))
```

For-hire and Private Truck have the highest percentages of shipments until
about 1,000 miles, where Parcel, USPS, or courier has the highest percentage of 
shipments.

## How does mode of transit change with shipment weight?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the percentages of packages transported via different modes, 
# by shipment weight
ggplot(data=cfs, aes(x=SHIPMT_WGHT)) +
  geom_freqpoly(aes(y=..count../sum(..count..), color=MODE)) + 
  scale_x_continuous(breaks=seq(0,6000,500), lim=c(500,3000))
```

Now subset the data to only include the predominany modes of transport.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot the percentages of packages transported by For-Hire Truck (04); 
# Private Truck (05); Air (including truck and water) (11); or Parcel, USPS, or
# courier (14)
ggplot(data=subset(cfs, MODE==04 | MODE==5 | MODE==11 | MODE==14), aes(x=SHIPMT_WGHT)) +
  geom_freqpoly(aes(y=..count../sum(..count..), color=MODE)) + 
  scale_x_continuous(breaks=seq(0,6000,200), lim=c(0,2000))
```

Parcel, USPS, or Courier has the highest percentage of shipments up to about 
70 pounds, and then falls precipitously to practically none of the shipments. 
For-Hire Truck and Private Truck have the highest percentage of shipments that
weigh greater than 70 pounds. Air Travel has some packages up to about 70
pounds, and then falls to a very small percentage of packages.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The main findings relate to the mode of transport used:

* Shipment Distance: For-hire and Private Truck have the highest percentages of 
shipments up about 1,000 miles, where Parcel, USPS, or courier has the highest 
percentage of shipments.
* Shipment Weight: Parcel, USPS, or Courier has the highest percentage of 
shipments up to about 70 pounds, and then falls precipitously to practically 
none of the shipments. For-Hire Truck and Private Truck have the highest 
percentage of shipments that weighh greater than 70 pounds. Air Travel has some 
packages up to about 70 pounds, and then falls to a very small percentage of 
packages.

### Were there any interesting or surprising interactions between features?
I was surprised the ratio between great circle distance and routed distance
behaved so erratically with exports. This might relate to the location where
shipments can legally enter each country.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# Plot a histogram of shipment value
ggplot(data=cfs, aes(SHIPMT_VALUE)) + 
  geom_histogram(binwidth=1000) + 
  scale_x_continuous(lim=c(0,10000000), breaks=seq(0,10000000,1000000)) +
  scale_y_continuous(lim=c(0,100)) +
  labs(title='Distribution of Shipment Value', x='Shipment Value ($)', y='Count')

head(arrange(cfs, desc(SHIPMT_VALUE)), 1)
```

### Description One
The distribution of shipment value is a right-skewed long-tailed distribution
with the majority of shipments falling below $1,000. The maximum valued shipment
of $521,277,800  was a 91,990 pound 'Transportation Equipment Manufacturing' 
shipment delivered 7 miles by Air from Dallas-Fortworth to Dallas-Fortworth. 

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# Plot the percentages of packages transported by predominant modes
ggplot(data=subset(cfs, MODE==04 | MODE==05 | MODE==14), aes(x=SHIPMT_DIST_ROUTED)) +
  geom_freqpoly(aes(y=(..count..)/sum(..count..), color=MODE)) + 
  scale_x_continuous(breaks=seq(0,6000,500), lim=c(0,6000), ) +
  labs(
    title='Shipment Distance vs. Percentage of Shipments by Predominant Modes',
    x='Shipment Distance (Miles)',
    y='Percentage of Shipments') +
  scale_color_hue(name='Mode of Transport',
                  labels = c('For-hire Truck', 
                             'Private Truck', 
                             'Parcel, USPS, or Courier'))
```

### Description Two
For-hire and Private Truck have the highest percentages of shipments for 
shipment distances up to about 1,000 miles, where Parcel, USPS, or courier has 
the highest percentage of shipments.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# Plot the percentages of packages transported by For-Hire Truck (04); 
# Private Truck (05); Air (including truck and water) (11); or Parcel, USPS, or
# courier (14)
ggplot(data=subset(cfs, MODE==04 | MODE==5 | MODE==11 | MODE==14), aes(x=SHIPMT_WGHT)) +
  geom_freqpoly(aes(y=..count../sum(..count..), color=MODE)) + 
  scale_x_continuous(breaks=seq(0,6000,200), lim=c(0,2000)) + 
  labs(
    title='Shipment Weight vs. Percentage of Shipments by Predominant Modes',
    x='Shipment Weight (Pounds)',
    y='Percentage of Shipments') +
  scale_color_hue(name='Mode of Transport',
                  labels = c('For-hire Truck', 
                           'Private Truck', 
                           'Air (incl truck & air',
                           'Parcel, USPS, or Courier'))
```

### Description Three
Parcel, USPS, or Courier has the highest percentage of shipments up to about 
70 pounds, and then falls precipitously to practically none of the shipments. 
For-Hire Truck and Private Truck have the highest percentage of shipments that
weigh greater than 70 pounds. Air Travel has a low percentage of
packages up to about 70 pounds, and then falls to a very small percentage of 
packages.

------

# Reflection

The main conclusions drawn from the exploration of the Commodity Flow Survey
dataset are:

* The shipments with the most weight and value tend to be in the agriculture, 
transportation, and energy industries.
* The heaviest shipments are transported mainly by rail, while the average 
shipments are transported mainly by road.
* Air, water, and rail are the modes of transit used for the longest shipment 
distances.
* For-hire and Private Truck have the highest percentages of 
shipments up about 1,000 miles, where Parcel, USPS, or courier has the highest 
percentage of shipments.
* Parcel, USPS, or Courier has the highest percentage of 
shipments up to about 70 pounds, and then falls precipitously to practically 
none of the shipments. For-Hire Truck and Private Truck have the highest 
percentage of shipments that weighh greater than 70 pounds. Air Travel has some 
packages up to about 70 pounds, and then falls to a very small percentage of 
packages.

Some challenges with the analysis relate to the large size (4,547,661 
observations). As a result of the size, a 10,000 observation sample was used
to explore different resource-heavy plots, such as the pairplot.

I would like to further explore different regressions that could be conducted
with different variables in the dataset. Can mode of transit be predicted by
any set of variables?